Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-11-02 manu-draft.Rmd
AA edits added

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L. Cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release of wastewater from Piney Point are abnormal relative to historic conditions in Tampa Bay.

*Key words*: algae, nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O). Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].  The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b]. Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain wastewater on site. There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks. Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Ecosystem management paradigms for estuaries are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.  The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued over the following months. These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.  The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of effluent release and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.

Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)). Details of the sampling methods are provided in supplement. Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point. This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.

@@

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the five month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

@@

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April Chl-a concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 and 3, respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.  Statistical comparisons between months for TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)). Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on Area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

@@

A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.  Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.  Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.  For seagrasses, turtle grass (*T. testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].  At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.

Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis sp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  *Dapis sp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change dramatically from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

@@

The increase in *K. brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021. In 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

>>>>

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site. Cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release of wastewater from Piney Point are abnormal relative to historic conditions in Tampa Bay.

*Key words*: eutrophication, macroalgae, nitrogen, phosphate mining, seagrass, Tampa Bay

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O). Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds capable of storing millions of liters of wastewater [@Burnett01].  The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b]. Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain wastewater on site. There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks. Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.  Nitrogen inputs from external sources are well understood as drivers of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks ([Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Department of Environmental Protection (FDEP).  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September. These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.  The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of effluent release and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.

Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)). Details of the sampling methods are provided in supplement. Additional samples for contaminants (e.g., heavy metals), invertebrates, and nekton were also collected, but they are not reported here in anticipation of future monitoring events.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference (see https://shiny.tbep.org/piney-point/ for a map of the matches).

@@

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the five month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates. For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

@@

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April Chl-a concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although bloom conditions were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.  Statistical comparisons between months for TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)). Further analsyis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

@@

A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.  Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.  Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence. For seagrasses, turtle grass (*T. testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].  At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.

A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis spp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  *Dapis spp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

@@

The increase in *K. brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021. Bloom concentrations in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations during June and July 2021 (Figure \@ref(fig:redtide)e).  In total, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa.  The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K. brevis* cell concentrations (Figure \@ref(fig:redtide)b).  Notably, all of the fish kill reports occured within a 1.5 month period, whereas red tide in the bay was observed for nearly four months. Increased fish kill reports in early July may coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f), which moved dead fish closer to populated nearshore areas, as noted earlier.  By August, Pinellas County and the city of St. Petersburg removed over 1600 metric tons of dead fish near public and private shoreline areas (K. Hammer Levy, Pinellas County, pers. comm. Aug. 2021).
